Imprecise Computing in Datapath Design: Precision Loss or Timing Violations

### 2013-04-10 TCAD2013/overclocking_tcad.tex
Modified CSA v4

\title{Title}

@@

@@

Conventionally a given frequency constraint is met by selecting word-lengths of each stage of the CSA in order to satisfy~(\ref{TimingConstraint_CSA}).
\begin{eqnarray}\label{TimingConstraint_CSA}
  \begin{matrix}
    d_i\leqslant\frac{1}{f_s} &, & \forall i\in[0,s-1]
  \end{matrix}
\end{eqnarray}
Under this situation, the maximum word-length of the most significant stage $n_{s-1}$ is given by~(\ref{WL_MSB_CSA}), where $b$ is determined by~(\ref{Max.length of carry chain}).
\begin{eqnarray}\label{WL_MSB_CSA}
  n_{s-1}\cdot\mu_c+\mu_{mux}=b\cdot\mu_c
\end{eqnarray}

Substitute~(\ref{WL_MSB_CSA}) into (\ref{CSA_WL}), we derive the representation of the word-length of CSA in terms of a given timing constraint, as presented in~(\ref{CSA_Timing}).
% we form the relationship between the word-length of CSA and RCA under a given timing constraint, as presented in~(\ref{CSA_Timing}).
\begin{eqnarray}\label{CSA_Timing}
  n_{CSA}=s\cdot b-\frac{\mu_{mux}}{\mu_{c}}\cdot\frac{(s+2)(s-1)}{2}
\end{eqnarray}

%Using this information, we verify our timing model with experimental results, which are obtained from post place and route simulations on Xilinx Virtex-6 FPGAs. Fig.~\ref{CSA Model Verification} illustrates both the modeled value and the experimental results of the maximum word-length of the 2-stage CSA for various operating frequencies. In the following experiments within this section, the input data are randomly sampled from a 16-bit data set, which follows uniform distribution. It can be seen that the the modeled value match well with the experimental results.
The ratio $\mu_{mux}/\mu_c$ can be computed from experiments. We perform post place-and-route simulations on the CSA with 2 stages using Xilinx Virtex-6 FPGA. The delay of the $i^{th}$ stage $d_i$ in~(\ref{CSA_SingleStageDelay}) is recorded with respect to different word-lengths of this given stage ($n_i$). Then the total word-length of CSA can be predicted through~(\ref{CSA_Timing}). In addition, the maximum word-length of the 2-stage CSA is obtained experimentally by increasing $n_{CSA}$ until errors are observed at the output. The comparison between the modeled value and the empirical results of $n_{CSA}$ is illustrated in Fig.~(\ref{CSA Model Verification}). It can be seen that our model match well with the experimental results.
%
%experimentally and compare our model with the empirical results of the maximum word-length of the 2stage CSA for various operating frequencies. It can be seen that the modeled value match well with the experimental results.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=3in]{./Figures/Model.eps}
  \caption{Comparison of the maximum word-length between the modeled value and the experimental results from FPGA simulations.}
  \label{CSA Model Verification}
\end{figure}

%It can be seen that our timing models match well the experimental results for both RCA and CSA, except that the modelled outcomes are slight conservative, especially at higher operating frequencies. This is because the model coefficients are obtained based on timing analysis, which is designed with safety margins to ensure correct functionality for diverse operating conditions. In addition, routing delay might be introduced during timing analysis such that the overall delay is enlarged, while our models consider logic delay only.

\subsection{Accuracy Benefits and Area Overhead in CSA}
As greater word-length corresponds to smaller truncation errors, we compare the maximum word-length of RCA and CSA with 2 stages and 3 stages over a range of operating frequencies, as presented in Fig.~\ref{CSA3stage Timing}. It can be seen from Fig.~\ref{CSA3stage Timing} that in comparison to RCA, CSA achieves greater word-length when frequency is initially increased. RCA only outperforms than CSA when very high frequency is applied. This is because at low frequencies, although the multiplexer delay limits the word-length of each stage in CSA when compared to RCA, the stage parallelism in CSA enables a greater word-length. However when frequency increases, the multiplexer delay becomes comparable to the delay of the carry chain, and this inhibits the benefits of parallelism. In addition, we can see that the word-length of 3-stage CSA is always greater than 2-stage CSA across the entire frequency domain.
\begin{figure}[tbp]
    \begin{minipage}[b]{.48\textwidth}
        \centering
        \subfigure[Maximum word-length.]{
            \includegraphics[width=3.3in]{./Figures/CSA3stage_Timing.eps}
            \label{CSA3stage Timing}
        }
        \subfigure[Hardware resource usage.]{
            \includegraphics[width=3.3in]{./Figures/CSA3stage_Area.eps}
            \label{CSA3stage Area}
        }
    \end{minipage}
    \centering
    \caption{Comparison between RCA and CSA across a variety of frequency values in terms of the maximum word-length of input signal and the area consumption.}
\end{figure}

However, the accuracy benefits brought by CSA comes at the cost of a large area overhead. Fig.~\ref{CSA3stage Area} depicts the number of Look-Up Tables (LUTs) in the FPGA used for all three structures. It can be seen that in order to meet a given frequency, the 3-stage CSA consumes $2.4\times\sim3.7\times$ area than RCA, while the 2-stage CSA requires $1.7\times\sim3.1\times$ extra area. This finding poses a question of which is the best adder structure for a specific area budget.
%whether CSA with higher stage numbers still offer better accuracy than RCA under a limited area budget.
%\begin{figure}[htbp]
%  \centering
%  \includegraphics[width=3.3in]{./Figures/CSA3stage_Area.eps}
%  \caption{Area overhead for RCA and CSA.}
%  \label{CSA3stage Area}
%\end{figure}

\subsection{Exploring Trade-offs Between Accuracy, Performance and Area}
In the conventional design scenario, the word-length of the RCA and CSA are limited by the following two factors:
\begin{itemize}
  \item The critical path of the adder;
  \item The available silicon area.
\end{itemize}
We compare the accuracy, performance and area trade-offs to an RCA where the maximum possible word-length is limited by the given area budget, and the timing constraints are allowed to be violated.
%To answer this question, the silicon area is incorporated into the trade-offs together with accuracy and performance. If the available hardware resources are limited, the full word-length of both CSA and RCA might not be implemented. The precision loss might results in large errors even at low frequencies. This is demonstrated by experiments where area constraints are applied besides the timing requirement. In the conventional design scenario, both of the constraints are met by reducing the word-length of the input signal. The errors at the outputs are recorded with the reference to the original word-length, which is 16 bits in the following experiments. In addition, we propose another circuit designing method where RCA is implemented with the maximum possible word-length under the given area budget, while the timing constraints are met by overclocking.

Suppose an algorithm designer would wish to create a circuit that can run at a given frequency with the minimum achievable output errors and resource usage. In this case, for a certain pair of $\{Area, Frequency\}$ constraint, the optimum design method is selected based on the following criteria:
%For instance, Fig.xxx to Fig.xxx illustrate error expectations of all the described design scenarios when setting the LUTs number to 45, 35, 25 and 15 respectively. The optimal design metric with the minimum error expectation is labelled for each area constraint. For a given frequency and area constraint, the optimum design is chosen based on the following metrics:
\begin{itemize}
  \item Design with the best accuracy at outputs is the optimum design;
  \item If multiple designs achieve the same accuracy, then the design with minimum area is the optimum design;
  \item If the accuracy and area are identical for multiple designs, they are all treated as the optimum design.
\end{itemize}

If we define the accuracy in terms of the error expectation at the output. The optimum design method with respect to a variety of operating frequencies and area consumptions is demonstrated in Fig.~\ref{Tradeoff}. From this figure several observations can be made. If the available area is large enough to implement a CSA in full precision, it will be the optimum design. This is expected from our earlier analysis in Fig~(\ref{CSA3stage Timing}). The 2-stage CSA is better than the 3-stage CSA when frequency is initially increased, as it consumes less area although both of them achieve the same error expectation. For a tighter area budget, only part of the CSA can be implemented, whereas the RCA still keeps full precision. In this case, area becomes the dominate factor and precision is lost for the CSA. In this case we see the RCA with overclocking is the optimum design method across almost the whole frequency domain. For more stringent area constraints, the word-length of RCA is also limited. This results in truncation error initially for all design scenarios. However, the RCA with overclocking is still .......
%Meanwhile, CSA with high stage numbers might not even be implemented at all.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=3.5in]{./Figures/Tradeoff.eps}
  \caption{Demonstration of the optimum design methodology which achieves the minimum error at outputs with respect to a variety of frequency and area constraints.}
  \label{Tradeoff}
\end{figure}

Similarly if the algorithm designer would expect the circuit to operate as fast as possible with the minimum area whilst a certain error budget can be tolerated, the optimum design methodology can be decided as illustrated in Fig.~(\ref{Tradeoff_Error}). In this situation, the error specifications are evaluated in terms of mean relative error (MRE), as presented in~(\ref{MRE}), where $E_{error}$ and $E_{out}$ refer to the mean value of error and the mean value of outputs, respectively.
\begin{eqnarray}\label{MRE}
  MRE=\left|\frac{E_{error}}{E_{out}}\right|\times 100\%
\end{eqnarray}

In our experiments, MRE is set ranging from $0.001\%$ to $50\%$. For a certain MRE, the design with the maximum operating frequencies is labeled as the optimum design. Moreover, the smallest design is the optimum one if multiple structures operate at the same frequency, with a certain area. Based on these criteria, the decision graph is depicted in Fig.~(\ref{Tradeoff_Error}).
%We also notice that for small frequencies, RCA with either overclocking or truncation of word-length of the input signal is the best design choice for any area constraint, since
\begin{figure}[htbp]
    \centering
    \includegraphics[width=3.5in]{./Figures/Tradeoff_Error.eps}
    \caption{Demonstration of the optimum design methodology which runs at the fastest frequency with respect to a variety of accuracy and area constraints.}
    \label{Tradeoff_Error}
\end{figure}

For a tight accuracy requirement, i.e. $MRE<0.005\%$, CSA serves as the optimum design choice with respect to large accessible area, as it intrinsically operates faster than RCA. Once again, when the area budget shrinks, the RCA performs best because the precision of the CSA is limited. Similarly to the previous section, we see that the overclocked RCA achieves the fastest operating frequencies with respect to most area constraints.

>>>>

\title{Imprecise Computing in Datapath Design: Precision Loss or Timing Violations}
%Imprecise Datapath Design: An Overclocking Approach
%Imprecise Computation Through Overclocking

@@

Conventionally under a given frequency constraint, the word-length of RCA is truncated by using $n_{RCA}=b-1$ bits, where $b$ is determined by~(\ref{Max.length of carry chain}). Similarly for CSA, the word-length of each stage should be selected in order to satisfy~(\ref{TimingConstraint_CSA}).
\begin{eqnarray}\label{TimingConstraint_CSA}
  \begin{matrix}
    d_i\leqslant\frac{1}{f_s} &, & \forall i\in[0,s-1]
  \end{matrix}
\end{eqnarray}

Hence for the same frequency requirement, we can form the relationship between the delay of the most significant stage of CSA and the delay of RCA, as given by~(\ref{WL_MSB_CSA}),
\begin{eqnarray}\label{WL_MSB_CSA}
    \mu_c\cdot (b-1) = n_{s-1}\cdot\mu_c+\mu_{mux}
\end{eqnarray}

Substitute~(\ref{WL_MSB_CSA}) into (\ref{CSA_WL}) to replace $n_{s-1}$, we derive the representation of the word-length of CSA in terms of a given timing constraint, as presented in~(\ref{CSA_Timing}).
% we form the relationship between the word-length of CSA and RCA under a given timing constraint, as presented in~(\ref{CSA_Timing}).
\begin{eqnarray}\label{CSA_Timing}
  n_{CSA}=s\cdot (b-1)-\frac{\mu_{mux}}{\mu_{c}}\cdot\frac{(s+2)(s-1)}{2}
\end{eqnarray}

%Using this information, we verify our timing model with experimental results, which are obtained from post place and route simulations on Xilinx Virtex-6 FPGAs. Fig.~\ref{CSA Model Verification} illustrates both the modeled value and the experimental results of the maximum word-length of the 2-stage CSA for various operating frequencies. In the following experiments within this section, the input data are randomly sampled from a 16-bit data set, which follows uniform distribution. It can be seen that the the modeled value match well with the experimental results.
The ratio $\mu_{mux}/\mu_c$ can be computed from experiments. We perform post place-and-route simulations on the CSA with 2 stages using Xilinx Virtex-6 FPGA. The delay of the $i^{th}$ stage $d_i$ in~(\ref{CSA_SingleStageDelay}) is recorded with respect to different word-lengths of this given stage ($n_i$). Then the total word-length of CSA can be predicted through~(\ref{CSA_Timing}). In addition, the maximum word-length of the 2-stage CSA is obtained experimentally by increasing $n_{CSA}$ until errors are observed at the output. The comparison between the modeled value and the empirical results of $n_{CSA}$ is illustrated in Fig.~(\ref{CSA Model Verification}). It can be seen that our model match well with the experimental results.
%
\begin{figure}[htbp]
  \centering
  \includegraphics[width=3in]{./Figures/Model.eps}
  \caption{Comparison of the maximum word-length between the modeled value and the experimental results from FPGA simulations.}
  \label{CSA Model Verification}
\end{figure}

\subsection{Accuracy Benefits and Area Overhead in CSA}
As greater word-length corresponds to smaller truncation errors, we compare the maximum word-length of RCA and CSA with 2 stages and 3 stages over a range of operating frequencies, as presented in Fig.~\ref{CSA3stage Timing}. It can be seen from Fig.~\ref{CSA3stage Timing} that in comparison to RCA, CSA achieves greater word-length when frequency is initially increased. RCA only outperforms than CSA when very high frequency is applied. This is because at low frequencies, although the multiplexer delay limits the word-length of each stage in CSA when compared to RCA, the stage parallelism in CSA enables a greater word-length. However when frequency increases, the multiplexer delay becomes comparable to the delay of the carry chain, and this inhibits the benefits of parallelism. In addition, we can see that the word-length of 3-stage CSA is always greater than 2-stage CSA across the entire frequency domain.
\begin{figure}[tbp]
    \begin{minipage}[b]{.48\textwidth}
        \centering
        \subfigure[Maximum word-length.]{
            \includegraphics[width=3.3in]{./Figures/CSA3stage_Timing.eps}
            \label{CSA3stage Timing}
        }
        \subfigure[Hardware resource usage.]{
            \includegraphics[width=3.3in]{./Figures/CSA3stage_Area.eps}
            \label{CSA3stage Area}
        }
    \end{minipage}
    \centering
    \caption{A comparison between RCA and CSA in terms of the maximum word-length of input signal and the area consumption.}
\end{figure}

However, the accuracy benefits brought by CSA comes at the cost of a large area overhead. Fig.~\ref{CSA3stage Area} depicts the number of Look-Up Tables (LUTs) in the FPGA used for all three structures. It can be seen that in order to meet a given frequency, the 3-stage CSA consumes $2.4\times\sim3.7\times$ area than RCA, while the 2-stage CSA requires $1.7\times\sim3.1\times$ extra area. This finding poses a question of which is the best adder structure for a specific area budget.

\subsection{Exploring Trade-offs Between Accuracy, Performance and Area}
In Section xxx, we have discussed two design scenarios when considering timing constraints. In this section, we expand our analysis by incorporating the silicon area as another evaluation metric, and investigate the accuracy, performance and area trade-offs for different adder structures. In the conventional design scenario, the word-length of RCA and CSA is limited by the given frequency constraint, or/and the available hardware resources. The precision loss potentially generates large errors even without timing violations. However in the new design scenario, we use RCA with the maximum possible word-length under the given area budget, and the timing constraints are allowed to be violated. This process might result in timing errors as well as truncation errors due to area limitation. We compare these two scenarios with different design goals, with the aim of targeting the optimum design methodology under each situation.

%To answer this question, the silicon area is incorporated into the trade-offs together with accuracy and performance. If the available hardware resources are limited, the full word-length of both CSA and RCA might not be implemented. The precision loss might results in large errors even at low frequencies. This is demonstrated by experiments where area constraints are applied besides the timing requirement. In the conventional design scenario, both of the constraints are met by reducing the word-length of the input signal. The errors at the outputs are recorded with the reference to the original word-length, which is 16 bits in the following experiments. In addition, we propose another circuit designing method where RCA is implemented with the maximum possible word-length under the given area budget, while the timing constraints are met by overclocking.

First, suppose the designer would wish to create a circuit that can run at a given frequency with the minimum achievable output errors and resource usage. That is, a pair of $\{Area, Frequency\}$ constraint is applied. In this case, the optimum design method is selected based on the following criteria:
%For instance, Fig.xxx to Fig.xxx illustrate error expectations of all the described design scenarios when setting the LUTs number to 45, 35, 25 and 15 respectively. The optimal design metric with the minimum error expectation is labelled for each area constraint. For a given frequency and area constraint, the optimum design is chosen based on the following metrics:
\begin{itemize}
  \item Design with the best accuracy at outputs is the optimum design;
  \item If multiple designs achieve the same accuracy, then the design with minimum area is the optimum design;
  \item If the accuracy and area are identical for multiple designs, they are all treated as the optimum design.
\end{itemize}

For instance let the available LUTs number be 25, the accuracy with respect to different operating frequencies for both design scenarios are presented in Fig.~(\ref{CSA_LUT25}), on which the optimum design metric is labeled. Note that the accuracy is defined in terms of the error expectation at the output.
\begin{figure}[ht]
  \centering
  \includegraphics[width=3.3in]{./Figures/Error_LUT25.eps}
  \caption{A comparison between two design scenarios when the number of LUT is set to 25. The RCA and CSA are investigated in the conventional scenario, while the RCA is explored in the proposed new scenario. The results are obtained from post place-and-route simulations on Xilinx Virtex FPGAs.}
  \label{CSA_LUT25}
\end{figure}

We first notice that for all frequency values, the overclocked RCA achieves smaller error expectation than the RCA with truncated operand word-lengths, as predicted by the models in Section xxx. It can also be observed that CSA cannot be implemented with full word-length due to the limited area budget, and this leads to large truncation errors initially for CSA.

We then perform the similar experiments with a variety of area constraints. The optimum design method with respect to different operating frequencies and area consumptions is demonstrated in Fig.~\ref{Fig_CSA_Tradeoff}. From this figure, several observations can be made. If the available area is large enough to implement a CSA in full precision, it will be the optimum design. This is expected from our earlier analysis in Fig.~\ref{CSA3stage Timing}. The 2-stage CSA is better than the 3-stage CSA when frequency is initially increased, as it consumes less area although both of them achieve the same error expectation. For a tighter area budget, only part of the CSA can be implemented, whereas the RCA still keeps full precision. In this case, area becomes the dominate factor and precision is lost for the CSA. We see the RCA with overclocking is the optimum design method across almost the whole frequency domain. For a more stringent area constraint, the word-length of RCA is also limited. This results in truncation errors initially for all design scenarios. However, the RCA with overclocking can still be employed as the optimum design, because it loses less precision than CSA under the same area constraint.
\begin{figure}[t]
  \centering
  \includegraphics[width=3.5in]{./Figures/Tradeoff.eps}
  \caption{A demonstration of the optimum design methodology which achieves the minimum error at outputs with respect to a variety of frequency and area constraints.}
  \label{Fig_CSA_Tradeoff}
\end{figure}

Similarly if the design goal is to operate the circuit as fast as possible with the minimum area whilst a certain error budget can be tolerated, the optimum design methodology can be decided as illustrated in Fig.~\ref{Fig_CSA_Tradeoff_Error}. In this situation, the error specifications are evaluated in terms of the mean relative error (MRE), as presented in~(\ref{MRE}), where $E_{error}$ and $E_{out}$ refer to the mean value of error and the mean value of outputs, respectively.
\begin{eqnarray}\label{MRE}
  MRE=\left|\frac{E_{error}}{E_{out}}\right|\times 100\%
\end{eqnarray}

In our experiments, MRE is set ranging from $0.001\%$ to $50\%$. For a certain MRE, the design with the maximum operating frequencies is selected as the optimum design. Moreover, the smallest design is the optimum one if multiple structures operate at the same frequency, with a certain area requirement. Fig.~\ref{Fig_CSA_Tradeoff_Error} can thus be obtained based on these criteria.
%Based on these criteria, the decision graph is depicted in
%We also notice that for small frequencies, RCA with either overclocking or truncation of word-length of the input signal is the best design choice for any area constraint, since
\begin{figure}[t]
    \centering
    \includegraphics[width=3.5in]{./Figures/Tradeoff_Error.eps}
    \caption{A demonstration of the optimum design methodology which runs at the fastest frequency with respect to a variety of accuracy and area constraints.}
    \label{Fig_CSA_Tradeoff_Error}
\end{figure}

For a tight accuracy requirement, i.e. $MRE<0.005\%$, CSA serves as the optimum design choice with respect to large accessible area, as it intrinsically operates faster than RCA. Once again, when the area budget shrinks, the RCA performs best because the precision of the CSA is limited. Similarly to the previous section, we see that the overclocked RCA achieves the fastest operating frequencies under most area constraints when the accuracy requirement is released.

To sum up, it can be seen that for both design goals, CSA is the best option only when there is enough hardware resources. However this circumstance is unlikely to happen, especially with the continued technology scaling nowadays. Due to this reason, in the following sections we will focus on the situation where the hardware resources are limited, and therefore RCA is the design option in our following analysis and experiments.

@@
